# Visualizing data with `ggplot2

## Introduction
Before we actually start to do stuff, we should load some libraries. 

```{r, message = FALSE}
rm(list=ls())         # clean workspace
library('ggplot2')    # graphics library
library('tibble')     # nice dataframes
```

### The dataset

While we are at it, let us also load a dataset which is included in the __ggplot__ library. 
```{r}
diamonds_df = as_data_frame(diamonds)
head(diamonds_df) # look at the head of the dataset
```
Let's briefly look at this dataset. It includes the prices and features  of around 54 000 round cut diamonds. It contains the following variables (Hint: you can also look up the description of the dataset using `?diamonds` once you loaded `ggplot2`): 

+ __price__: price in US dollars 
+ __carat__: weight of the diamond
+ __cut__: quality of the cut (Fair, Good, Very Good, Premium, Ideal)
+ __color__: diamond colour from J (worst) to D (best)
+ __clarity__: clarity of the diamond (l1, SI1, SI2, VS1, VS2, VVS1, VVS2, IF)
+ __x__: length in mm
+ __y__: width in mm
+ __z__: depth in mm
+ __depth__: total depth percentage
+ __table__: width of top of diamond relative to widest point

We will use this dataset in the following to learn the basics of `ggplot2`, a graphics package for R which can be used1 to create beautiful graphics. 

###  Basic example: an ordinary scatterplot
We will start to build up the logic of the `ggplot2` command step by step. No matter which graphics you want to draw, you have to start as follows:

```{r}
ggplot(data = diamonds_df)
```

As you can see, if you execute this command, this opens an empty graphics windows and nothing else happens. This make sense. So far we have only told `ggplot` which data we want to use for the plot. 

To actually see something we have to add a layer. Say we want to produce a scatter plot that shows us the association between the variables `carat` and `price`. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat , y = price))
```

What have we done here? We have added a new layer to our empty plot. The function `geom_point` specifies that we want to add a scatterplot layer. The `mapping` argument defines how data are mapped into the visual properties of the graph. This argument is always combined with the `aes` function which is a shorthand for `aesthetics`. 

In our case, we have specified to map the carat of the diamond to the x-axis and the price of the diamond to the y-axis. There seems to be a positive relationship between `carat` and `price` although it is noisier for higher carat levels. 

##  Aesthetics
###  Aesthetic mappings 
Of course, we can do more by adding further variables to our plot. Say we want to color the dots according to the clarity category. For this, we have to add the `color` argument to the `aes`function. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, color = clarity))
```

We have mapped  the `clarity` variable to the `color` aesthetic and this gives us a clear pattern: For a given weight of the diamond, diamonds with higher clarity are more expensive. There are also other aesthetics and we could have mapped the variable into each of them. Some aesthetics are:

+ size of the dots with `size`
+ transparency of the dots with `alpha`
+ shape of dots with `shape`

Let's try it out: 
```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, alpha = clarity))
```

Pretty cool, eh? 

You can also map more than simple variables to aesthetics. Here we produce a boolean which is `TRUE` or `FALSE` depending on whether clarity is above a certain threshold. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, color = clarity > 'VS2'))
```


### Manually fixing aesthetics 
So, what if you just want to manually change an aesthetic without mapping a variable to it? For example, we might have a strong affinity to the color blue and prefer it to the depressing black of our graphs. To do that, we can just pass the name of the color as a string to the `color` argument and pass it outside of the `aes` function.

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price), color = "blue")
```

You can do similar things with the other aesthetics: 

+ for `size`: size in mm
+ for `shape`: shape of a point as a number
+ for `alpha`: value between 0 and 1  

## Facets: Subplots for different variable values
What if we want do not map all variables into aesthetics of one graph but want to have one graph for each value of a variable? For our dataset, we could for example be interested in how carat is associated with price for different values of the variable `cut`. For this, we need to split our plot into different facets (i.e. subplots) using the `facet_wrap` function.

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price)) + 
  facet_wrap(~ cut, nrow = 2)
```

As you can see, for each value of the `cut` variable, we get a separate subplot. Note that you should pass a discrete variable to the `facet_wrap` function. The `nrow` argument specified into how many rows the plots are organized. We can also cross-tabulate subplots for different variables with the `facet_grid` function. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price)) + 
  facet_grid(clarity ~ cut)
```

You can also use `facet_grid` to basically replicate the plot generated with `facet_wrap`. 
```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price)) + 
  facet_grid(. ~ cut)
```

## Geoms
### What kind of plot do you have in mind? The use of geoms. 
So far we have just produced a scatterplot. We can of course also create other kinds of plots using different geometrical objects or `geoms`. For example, we could fit a smooth line to the data with `geom_smooth`. 

```{r}
ggplot(data = diamonds_df) + 
  geom_smooth(mapping = aes(x = carat, y = price))
```

Depending on the geometrical object, different aesthetics can be used. To see which aesthetics are available type `?geom_smooth`. Moreover, you can also see which additional arguments are available and which default values are set. In the following, we will change the confidence intervals and map the variable `cut` to the `color` aesthetic. 

```{r}
ggplot(data = diamonds_df) + 
  geom_smooth(mapping = aes(x = carat, y = price, color = cut), level = 0.9)
```

### Multiple geoms in one plot
We can also put multiple geoms in one graph by creating multiple layers. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price), size = 0.2, alpha = 0.5) +
  geom_smooth(mapping = aes(x = carat, y = price))
```

As you can see, we have simply added another layer to our plot using the `+` operator. Still, this is a little redundant since we pass the same values for the same aesthetics two times. We can improve on this a little, by putting common aesthetics in the `ggplot` command.  

```{r}
ggplot(data = diamonds_df, mapping = aes(x = carat, y = price)) + 
  geom_point(size = 0.2, alpha = 0.5) + 
  geom_smooth()
```

It's the same graph but with less typing! Note that I could also have put additional aesthetics or overwritten the global ones in the `geom_point` and `geom_smooth` functions. 

There are tons of geoms from which you can choose your preferred one. Instead of explaining them all, I have distributed a pdf document with a very helpful cheat sheet. Feel free to play around with them!

## Statistical tranformations with `stats`
When you use `ggplot`, there is always a statistical tranformation - a stat - of the data running in the background. Let us look at the following two plots. The first should be familiar by now. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price))
```

What happens in the background? We have plotted the raw data for the variables `carat` and `price` on the x- and y-axis. Nevertheless, in the background `ggplot` used a statistal transformation. It took the data from our dataframe, applied the `identity` transformation, i.e. nothing is changed, and passed the transformed data to `ggplot`. Not so exiting, right? Let's look at a second plot, a histogram for the variable `clarity`.

```{r}
ggplot(data = diamonds_df) + 
  geom_bar(mapping = aes(x = clarity))
```

The `x` axis plots the categories for clarity. The variable `count` is plotted on the y-axis. But you will not find the variable `count` in our dataset. `Ggplot` created a new dataframe with the variable `clarity` and the variable `count` for each category and plotted this to the figure. This is an example of the `count` stat!

At first, stats might not seem like an important topic because `ggplot` usually uses sensible statistical transformations. However, sometimes you want to override the default stat. Consider the following dataset now.

```{r}
test_df <- tibble(
  colors = c("green", "blue", "red", "black", "white"),
  numbers = c(20, 40, 50, 66, 8)
)
head(test_df)
```

What happens if we naively use the geom for histograms? 

```{r}
ggplot(data = test_df) + 
  geom_bar(mapping = aes(x = colors))
```

Well, this is not what we wanted. `Ggplot` did not realize that we already have summary statistics in this dataframe, therefore each color category it counted to appear once. We want the geom to use the data in the `numbers` variable for the y-axis. We need to use a different stat. 
```{r}
ggplot(data = test_df) + 
  geom_bar(mapping = aes(x = colors, y = numbers), stat = "identity")
```
Ah, this is much better. 

A second case where you might want to think about stats is when you want to override the default mapping from transformed values to aesthetics. Let's take a look at the help page for `geom_bar` by typing `?geom_bar`. If you scroll down you will see that `geom_bar` computes two variables by default, `count` and `prop`. Let's say we want ggplot to use proportions, i.e. `prop` instead of counts for the y-axis. This can be accomplished as follows: 

```{r}
ggplot(data = diamonds_df) + 
  geom_bar(mapping = aes(x = clarity, y = ..prop.., group = 1))
```

Here, we are telling `ggplot` to use the default statistical tranformation, but use the variable `prop` that is generated during the transformation for the y-axis.  

Again, there are tons of statistical transformations in `ggplot`.

## Positions 
Ok, now it is time to talk about position adjustments.  What happens in our histograms when we map a variable to the `fill` aesthetic?
```{r}
ggplot(data = diamonds_df) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```
This is cool, right? The bars for each `clarity` category are stacked upon each other for each `cut` category. This is because `ggplot` automatically uses the `stack` position adjustement when drawing `geom_bar`. There are more options however. 

```{r}
ggplot(data = diamonds_df) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "identity", alpha = 0.2)
```

You can see that the bars now all start at $y = 0$ and are not stacked upon each other. Let's look at another example: If you instead use the `fill` position, bars are stacked upon each other, but each stack of bars is forced to have the same height. 

```{r}
ggplot(data = diamonds_df) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

Third, you can also use the `dodge` position to make a useful change to the bar plot. This places the bars next to each other. 

```{r}
ggplot(data = diamonds_df) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

Finally, there is also a useful position adjustment for scatter plots. When points are plotted in the scatter plot their coordinates are rounded to be placed on the grid. Then, it can happen that many points overlap each other. To remidy this, you can use the `jitter` position adjustement which adds a small amount of random noise to each point. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price), position = "jitter")
```
Of course there is more to learn, but for that please be referred to the help pages by typing for example `? position_jitter`.

## Coordinate systems
The default coordinate system used by `ggplot` is the Cartesian one (`coord_cartesian`). I will only briefly introduce some other useful commands here. Rest assured that there is again much more to say here.

You can quickly flip axes using the `coord_flip` command. Let's just do this to one of our bar charts. 
```{r}
ggplot(data = diamonds_df) + 
  geom_bar(mapping = aes(x = cut)) + 
  coord_flip()
```

With the `coord_trans` function you plot data to a Cartesian coordinate system, where axes are transformed by a function. 
```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price)) + 
  coord_trans(x = "log", y = "log")
```

Sometimes you want to have a fixed aspect ratio for your coordinate system, you can use the `coord_fixed` command to create a Cartesian coordinate system with a fixed aspect ratio between `x` and `y` units. 

## Labeling 
So far, we looked at the construction of graphs but did not change labels. We can do this using the `labs` function. In this case, we add a title and labels for the x-axis and y-axis as well as the legend. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, color = clarity)) + 
  labs(title = "Relationship between weight and price of diamonds", 
       x = "Weight in carat", 
       y = "Price in dollars", 
       color = "Clarity")
```

## Scales
Apart from changing the labels you can also change the scales associated with your graph. Scales control how your data are mapped to the graph. These scales are usually applied automatically when you create a graph. Let's take our graph from the previous section. `Ggplot2` adds the following features in the background.  

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, color = clarity)) + 
  labs(title = "Relationship between weight and price of diamonds", 
       x = "Weight in carat", 
       y = "Price in dollars", 
       color = "Clarity") +
  scale_x_continuous() + 
  scale_y_continuous() + 
  scale_color_discrete()
```

As you can see, `ggplot2` automatically adds sensible scales. The first word after the `scale_` gives you the name of the aesthetic which is then followed by the name of the scale. Sometimes you can change graphs by passing arguments to the scale functions, sometimes you completely replace them by invoking different scale functions. We will talk here specifically about three things that we can tweak using scale functions: axis ticks, legend labels and color schemes. 

To change the axis ticks, you have to pass the `breaks` argument with the desired breaks to the scales. 
```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, color = clarity)) + 
  labs(title = "Relationship between weight and price of diamonds", 
       x = "Weight in carat", 
       y = "Price in dollars", 
       color = "Clarity") +
  scale_x_continuous(breaks = seq(0, 5, 0.5)) + 
  scale_y_continuous() + 
  scale_color_discrete()
```
You can also change the labels by passing the `labels` argument to the scale with the vector of desired labels. 

Next, we want to change the labels for our legend. These are related to the color aesthetic. We can just pass the `labels` argument with the vector of desired labels. In this example, I just pass a vector of numbers. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, color = clarity), size = 0.2) + 
  labs(title = "Relationship between weight and price of diamonds", 
       x = "Weight in carat", 
       y = "Price in dollars", 
       color = "Clarity") +
  scale_x_continuous() + 
  scale_y_continuous() + 
  scale_color_discrete(labels = seq(1, 8))
```

Finally, what if you want to change the color of our graph? It is obviously related to the color aesthetic. In this case, we don't have to pass an argument to the scale function but instead have to completely replace the scale. 
```{r}
# brewer scales at http://colorbrewer2.org
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, color = clarity), size = 0.2) + 
  labs(title = "Relationship between weight and price of diamonds", 
       x = "Weight in carat", 
       y = "Price in dollars", 
       color = "Clarity") +
  scale_x_continuous() + 
  scale_y_continuous() + 
  scale_color_brewer(palette = "Reds")
```
If you want to manually set colors, you can do so with the `scale_color_manual` function. 

## Themes
You can also change the complete non-data elements of your graph by applying the `theme` function. You have the following themes available: 

* `theme_bw`: white background with grid lines
* `theme_classic`: classic theme; axes but not grid lines
* `theme_dark`: dark background for contrast
* `theme_gray`: default theme with grey background
* `theme_light`: light axes and grid lines
* `theme_linedraw`: only black lines
* `theme_minimal`: minimal theme, no background
* `theme_void`: empty theme, only geoms are visible

There is also a package called `ggthemes` which gives you a ton of other templates. In the following, we just switch our graph to the classic theme. 

```{r}
ggplot(data = diamonds_df) + 
  geom_point(mapping = aes(x = carat, y = price, color = clarity), size = 0.2) + 
  labs(title = "Relationship between weight and price of diamonds", 
       x = "Weight in carat", 
       y = "Price in dollars", 
       color = "Clarity") +
  scale_color_brewer(palette = "Reds") +
  theme_classic()
```

## Saving graphs
We have not yet talked about how to save a plot to disk. This can be done using the `ggsave` function. Look up the help file for `ggsave` to see into which formats you can export your plot.

## Sources {-}
The exposition here is heavily inspired by the notes for a new book on R data science by Garrett Grolemund and Hadley Wickham. You can find detailed outlines here: http://r4ds.had.co.nz. The cheat sheet for `ggplot2` is from the official RStudio website: https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf. Finally, if you want to know more about the details of `ggplot2` there is a complete book on it by Hadley Wickham: https://www.amazon.de/ggplot2-Elegant-Graphics-Data-Analysis/dp/331924275X/ref=sr_1_1?ie=UTF8&qid=1474400904&sr=8-1&keywords=ggplot2. 