# (PART) Regression Modelling with R {-}

# Linear Regression Essentials with `R`

## Libraries
```{r, warning=FALSE, message=FALSE}
library(haven)      # reading stata data
library(dplyr)      # data manipulation
library(tibble)     # nicer dataframes
library(stargazer)  # tables
library(ggplot2)    # graphs
```

## Loading the data

```{r}
mrw_df = read_dta('data/mrw.dta')
head(mrw_df)
```

We can cutify the output a little by using the `kable` function.
```{r}
knitr::kable(head(mrw_df))
```

As you can see, we have `r ncol(mrw_df)` variables and `r nrow(mrw_df)` observations. 

We have the following variables: 

* __number__: a country identifier between 1 and 121 country country name (a string variable)
* __country__: the name of the country
* __n__: a dummy variable equal to one if the country is included in the non-oil sample
* __i__: a dummy variable equal to one if the country is included in the intermediate sample
* __o__: a dummy variable equal to one if the country is included in the oecd sample
* __rgdpw60__: real GDP per working age population in 1960
* __rgdpw85__: real GDP per working age population in 1985
* __gdpgrowth__: average annual growth rate of real GDP per working age population between 1960 and 1985 
* __popgrowth__: average annual growth rate of the working age population
between 1960 and 1985
* __i_y__: real investment as a share of real GDP, averaged over the period 1960-85
* __school__: % of working age population in secondary school

## Meaningful names
The first thing we should do is probably to give these variables more meaningful names in order to escape the 90s charme conveyed by them. 

```{r}
mrw_df = mrw_df %>% 
  rename(non_oil = n, 
         oecd = o,
         intermediate = i,
         gdp_60 = rgdpw60,
         gdp_85 = rgdpw85,
         gdp_growth_60_85 = gdpgrowth,
         pop_growth_60_85 = popgrowth,
         inv_gdp = i_y,
         school = school)
```

## Create variables for estimation
In order to follow the estimation, we will need to create some additional variables: 

* The logs of the GDP per working age pop. in 1985 and 1960.
* The investment to GDP ratio has to be converted to lie between 0 and 1. Also we need the log of it. 
* We have to create the `ndg` variable which is assumed to be population growth (0 - 1) + 0.05. Again, we need the log of it. 
* We want to use the log of the schooling rate (again first divided by 100). 
* Finally, and just for consistency, we should convert our sample dummies to factors. 


```{r}
# log gdp 
mrw_df = mrw_df %>% 
  mutate(ln_gdp_85 = log(gdp_85),
         ln_gdp_60 = log(gdp_60),
         ln_inv_gdp = log(inv_gdp/100),
         non_oil = factor(non_oil),
         intermediate = factor(intermediate),
         oecd = factor(oecd),
         ln_ndg = log(pop_growth_60_85/100 + 0.05),
         ln_school = log(school/100)) %>% 
  select(country, ln_gdp_85, ln_gdp_60, ln_inv_gdp, 
         non_oil, intermediate, oecd,
         ln_ndg, ln_school, gdp_growth_60_85)
head(mrw_df)
```

## Summary statistics

Maybe, we would like to have summary statistics for our dataframe. For that, we need `summary`.

```{r}
summary(mrw_df)
```

## Create three samples

```{r}
mrw_oecd = mrw_df %>% filter(oecd == 1)
mrw_int = mrw_df %>% filter(intermediate == 1)
mrw_non_oil = mrw_df %>% filter(non_oil == 1)
```

## Run the estimation for Table 1 in MRW (1992)
To run a linear model we need the `lm` command. 

```{r}
m_non_oil = lm(ln_gdp_85 ~ 1 + ln_inv_gdp + ln_ndg, data = mrw_non_oil)
m_int = lm(ln_gdp_85 ~ 1 + ln_inv_gdp + ln_ndg, data = mrw_int)
m_oecd = lm(ln_gdp_85 ~ 1 + ln_inv_gdp + ln_ndg, data = mrw_oecd)
```

To get nicely formatted results, we can use the `summary` command: 
```{r}
summary(m_non_oil)
```


## Show the results in a table
```{r, results='asis'}
stargazer(m_non_oil, m_int, m_oecd, type = "latex")
```

```{r, results='asis'}
stargazer(m_non_oil, m_int, m_oecd, type = "latex",
          column.labels = c("Non-Oil", 
                            "Intermediate", 
                            "OECD"),
          covariate.labels = c("$\\log(\\frac{I}{GDP})$", 
                               "$\\log(n+\\delta+g)$", 
                               "Constant"), 
          dep.var.labels = "Log(GDP) 1985",
          omit.stat = c("f", 
                        "rsq", 
                        "ser"),
          title = "Replication of (part of) Table 1 in Mankiw, Romer, and Weil (1992)",
          style = "qje")
```

## Robust standard errors

In economics, we often would like to have robust standard errors. To look at how we see them. Let's go back to an example. 


```{r}
lm_example = lm(ln_gdp_85 ~ 1 + ln_inv_gdp + ln_ndg, data = mrw_non_oil)
```

```{r}
library(sandwich) # for robust standard errors
library(lmtest)   # to nicely summarize the results

lm_robust = coeftest(lm_example, vcov = vcovHC(lm_example, "HC1"))
print(lm_robust)
```
Since we do not want to type this every time. We should write a short function that takes a linear model and returns the robust summary of it. 

```{r}

# needs sandwich and lmtest
print_robust = function(lm_model) {
  results_robust = coeftest(lm_model, vcov = vcovHC(lm_model, "HC1"))
  print(results_robust)
}

print_robust(lm_example)
```

Now, unfortunately the `coeftest` function does not return an object that is easily transferred to a stargazer table. Thus, we will have to write another function. 
```{r}
# needs sandwich
compute_rob_se = function(lm_model) {
  vcov = vcovHC(lm_example, "HC1")
  se = sqrt(diag(vcov))
}
```
This makes our life somewhat easier. No, in order to compare the standard errors we could do the following. 
```{r, results = 'asis'}
# run the model 
lm_example = lm(ln_gdp_85 ~ 1 + ln_inv_gdp + ln_ndg, data = mrw_non_oil)

# obtain the robust ses
rob_se = compute_rob_se(lm_example)

stargazer(lm_example, lm_example,
          se = list(NULL, rob_se))
```

## Some Graphs

```{r, fig.width = 6}
ggplot(mrw_non_oil) +
  geom_point(aes(x = ln_gdp_60, y = gdp_growth_60_85)) +
  labs(x = "Log output per working-age adult in 1960",
       y = "Growth rate: 1960-85", 
       title = "Unconditional Convergence") +
  theme_bw()
```

Let's try to get the residuals.

```{r, fig.width = 6}
lm_y = lm(gdp_growth_60_85 ~ 1+ ln_inv_gdp + ln_ndg + ln_school, data = mrw_non_oil)
lm_x = lm(ln_gdp_60 ~ 1+ ln_inv_gdp + ln_ndg + ln_school, data = mrw_non_oil)
y_res = lm_y$residuals
x_res = lm_x$residuals

graph_tibble = tibble(
  y = y_res, 
  x = x_res)

ggplot(graph_tibble) +
  geom_point(aes(x, y)) + 
  labs(x = "Res. X",
       y = "Res. Y") + 
  theme_bw()
```
