# Linear Models with Fixed Effects

## Libraries
```{r, message = FALSE}
library(readxl)         # read excel files
library(tibble)         # cuter dataframes
library(dplyr)          # data manipulation
library(ggplot2)        # graphs
library(lfe)            # fixed effects models
library(stargazer)      # nice tables
library(ggrepel)        # better graph labeling
library(lmtest)         # for coeftest function
library(multiwayvcov)   # (multiway) clustered standard errors
library(AER)            # instrumental variables
library(ivpack)         # robust standard errors for ivreg
```

## Review of Plotting: Recreating Figure 1
```{r}
# read data for first figure
ajry_f1 = read_xls("data/ajry.xls", 
                   sheet = "F1") %>% 
  rename(log_gdp_pc = lrgdpch, 
         freedom_house = fhpolrigaug)

ggplot(ajry_f1, aes(x = log_gdp_pc, y = freedom_house)) +
  geom_point(size = 0.5) +
  geom_text(aes(label = code), size = 2, hjust = 0, vjust = 0) +
  geom_smooth(method = "lm", color = "black", size = 0.5, alpha = 0.2) +
  labs(x = "Log GDP per Capita (1990-1999)",
       y = "FH Measure of Democracy (1990-1999)") +
  theme_bw()

```
That is not bad, but we cannot read half of the label. Let's try again, with the `ggrepel` package. 

```{r}
ggplot(ajry_f1, aes(x = log_gdp_pc, y = freedom_house)) +
  geom_point(size = 0.5) +
  geom_text_repel(aes(label = code), size = 2) +
  geom_smooth(method = "lm", color = "black", size = 0.5, alpha = 0.2) +
  labs(x = "Log GDP per Capita (1990-1999)",
       y = "FH Measure of Democracy (1990-1999)") +
  theme_bw()

rm(ajry_f1)
```

## Review of Plotting: Recreating Figure 2
```{r}
# read data for second figure
ajry_f2 = read_xls("data/ajry.xls", 
                   sheet = "F2") %>% 
  rename(freedom_house_change = s5fhpolrigaug,
         log_gdp_pc_change = s5lrgdpch)


ggplot(ajry_f2, aes(x = log_gdp_pc_change, y = freedom_house_change)) +
  geom_point(size = 0.5) + 
  geom_smooth(method = "lm", size = 0.5, alpha = 0.2) +
  geom_text_repel(aes(label = code), size = 2) +
  labs(x = "Change in GDP per Capita (1970-1995)",
       y = "Change in FH Measure of Democracy (1970-1995)") + 
  theme_bw()
```


## Loading the data for estimation
```{r}
ajry_df = read_xls("data/ajry.xls", 
                   sheet = 2) %>% 
  arrange(code_numeric, year_numeric) %>% 
  rename(log_gdp_pc = lrgdpch,
         freedom_house = fhpolrigaug)

# generate lagged variables
ajry_df = ajry_df %>% 
  group_by(code_numeric) %>% 
  mutate(lag_log_gdp_pc = lag(log_gdp_pc, order_by = year_numeric),
         lag_freedom_house = lag(freedom_house, order_by = year_numeric),
         lag2_nsave = lag(nsave, 2, order_by = year_numeric),
         lag_worldincome = lag(worldincome, order_by = year_numeric)) %>% 
  filter(sample == 1)
```

## Pooled OLS with Time Effects
```{r}
# pooled ols with lm 
pooled_est = lm(freedom_house ~ -1 + lag_freedom_house + lag_log_gdp_pc + 
             factor(year_numeric), data = ajry_df)

# standard errors clustered by country
vcov_country <- cluster.vcov(pooled_est, ajry_df$code_numeric)
coeftest(pooled_est, vcov_country)
```

## Fixed Effects with the `lm` function

```{r}
# pooled ols with lm 
fe_est = lm(freedom_house ~ -1 + lag_freedom_house + lag_log_gdp_pc + 
             factor(year_numeric) + factor(code_numeric), data = ajry_df)

# standard errors clustered by country
vcov_country <- cluster.vcov(fe_est, factor(ajry_df$code_numeric))
coeftest(fe_est, vcov_country)
```

## Pooled OLS and FE with the `lfe` package
```{r}
# pooled OLS
felm1 = felm(freedom_house ~ lag_freedom_house + lag_log_gdp_pc | year_numeric | 0 | code_numeric, 
             data = ajry_df)

# FE
felm2 = felm(freedom_house ~ lag_freedom_house + lag_log_gdp_pc | year_numeric + code_numeric | 0 | 
               code_numeric, data = ajry_df)
stargazer(felm1, felm2, type = 'text')
```
Notice that this command automatically spits out cluster-robust standard errors and passes them to `stargazer`. That is absolutely fantastic, right? For robust standard errors we have to work a bit. 

```{r}
# function to recover robust standard errors 
get_felm_robust_se = function(felm_result) {
  felm_summary = summary(felm_result, robust = TRUE)
  robust_se = felm_summary$coefficients[, 2]
  }
```

## Review of IV

```{r}
# Second Stage with ivreg, normal standard errors
iv_sav = ivreg(freedom_house ~ lag_freedom_house + lag_log_gdp_pc + factor(year_numeric) + 
        factor(code_numeric) | lag_freedom_house + lag2_nsave + factor(year_numeric) + 
        factor(code_numeric), data = ajry_df)
summary(iv_sav) 
```

## IV with `felm`

```{r}
# note the difference in the instrumental variable list.
summary(felm(freedom_house ~ lag_freedom_house | year_numeric + code_numeric | 
               (lag_log_gdp_pc ~ lag2_nsave) | code_numeric, data = ajry_df))
```
